---
title: 'Canonical Correlation Analysis & Reduced Rank Regression'
author: Casey Tirshfield
output: html_notebook
---

## The file $\texttt{m_logret_4auto.txt}$ contains the monthly log returns of four automobile manufacturers (General Motors Corp., Toyota Motor Corp., Ford Motor Co., and Honda Motor Co.) from January 1994 to June 2007. The file $\texttt{m_logret_4soft.txt}$ contains the monthly log returns of four application software companies (Adobe Systems Inc., Microsoft Corp., Oracle Corp., and SPSS Inc.) from January 1994 to June 2007.

```{r}
auto_df <- read.table('m_logret_4auto.txt', header=TRUE)
soft_df <- read.table('m_logret_4soft.txt', header=TRUE)

auto_df<- auto_df[,-1]
soft_df <- soft_df[,-1]
```

### (a) Perform a canonical correlation analysis for these two sets of returns. Give the first two estimated canonical variate pairs and the corresponding canonical correlations.

```{r}
# this code was inspired by https://stats.idre.ucla.edu/r/dae/canonical-correlation-analysis/
# determine correlations
canonical <- matcor(auto_df, soft_df)
print(canonical)

cc1 <- cc(auto_df, soft_df)

# display the canonical correlations
cc1$cor

# raw canonical coefficients
cc1[3:4]

# compute canonical loadings
cc2 <- comput(auto_df, soft_df, cc1)

# display canonical loadings
canon_var <- cc2[3:6]

# tests of canonical dimensions
ev <- (1 - cc1$cor^2)

n <- dim(auto_df)[1]
p <- length(auto_df)
q <- length(soft_df)
k <- min(p, q)
m <- n - 3/2 - (p + q)/2

w <- rev(cumprod(rev(ev)))

# initialize
d1 <- d2 <- f <- vector("numeric", k)

for (i in 1:k) {
    s <- sqrt((p^2 * q^2 - 4)/(p^2 + q^2 - 5))
    si <- 1/s
    d1[i] <- p * q
    d2[i] <- m * s - p * q/2 + 1
    r <- (1 - w[i]^si)/w[i]^si
    f[i] <- r * d2[i]/d1[i]
    p <- p - 1
    q <- q - 1
}

pv <- pf(f, d1, d2, lower.tail = FALSE)
(dmat <- cbind(WilksL = w, F = f, df1 = d1, df2 = d2, p = pv))

# standardized auto canonical coefficients diagonal matrix of auto sd's
s1 <- diag(sqrt(diag(cov(auto_df))))
s1 %*% cc1$xcoef

# standardized soft canonical coefficients diagonal matrix of soft sd's
s2 <- diag(sqrt(diag(cov(soft_df))))
s2 %*% cc1$ycoef

# this code was inspired by page 202 of the textbook
# first pair
alpha_pair1 <- canonical$Xcor %*% canon_var$corr.X.xscores[1,]
beta_pair1 <- canonical$Ycor %*% canon_var$corr.Y.xscores[1,]
print(alpha_pair1)
print(beta_pair1)

# second pair
alpha_pair2 <- canonical$Xcor %*% canon_var$corr.X.xscores[2,]
beta_pair2 <- canonical$Ycor %*% canon_var$corr.Y.xscores[2,]
print(alpha_pair2)
print(beta_pair2)
```

### (b) Perform reduced-rank regression of the log returns of automobile stocks on those of software company stocks, taking $rank(\mathbf{B}) = 2$ in $\mathbf{y}_k = \mathbf{Bx}_k+\epsilon_k, \quad k=1,\dots,n$ with response variable $\mathbf{y}_k=(y_{k1},\dots, y_{kq})^\top\in\mathbb{R}^q$, predictor variable $\mathbf{x}_k=(x_{k1},\dots,x_{kp})^\top\in\mathbb{R}^p$, and random error $\epsilon_k\in\mathbb{R}^q$ such that $\mathbb{E}(\epsilon_k)=\mathbf{0}$ and $Cov(\epsilon_k)=\mathbf{0}$ and $Cov(\epsilon_k)=\Sigma$.

```{r}
# calculate the regression residuals for reduced-rank regression and canonical variate analysis
residuals(soft_df, auto_df, type='identity', rank = 'full', k=0, plot=TRUE)

# fit the reduced-rank regression model
rrr(soft_df, auto_df, type = "identity", rank = '2', k = 0)
```

